---
title: "Affymetrix GeneChip Data Analyais with LIMMA"
author: "Asad"
date: "2023-05-13"
output: 
  pdf_document: 
    highlight: zenburn
---

## Call Libraries
```{r}
library(affy)
library(oligo)
library(GEOquery)
library(Biobase)
library(tidyr)
library(splitstackshape)
library(arrayQualityMetrics)
library(dplyr)
library(limma)
library(annotate)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(hgu133plus2.db)
library(EnhancedVolcano)
library(umap)
library(maptools)
setwd("D:/CancerData/Data")
```

## Read celFiles
```{r}
celFiles <- list.celfiles()
gset <- read.celfiles(celFiles)
```

## Check if the expression ranges between 0-16 which means log2 transformed data
```{r}
head(exprs(gset), 10)
```

## Check sample distribution and probe intensity plot before normalization
```{r}
boxplot(gset, xlab='Samples', ylab='log2 Expression', col='Red',
        main= 'Average Expression Before Normalization') 
hist(gset, lwd=2,xlab='Log intensity',
     main="Signal Densities Before Normalization")
```

#Generate pseudo-image of chip intensity for individual sample (CEL file)
#image(affyRaw[,1])

#Perform normalization
#RMA Normalization
```{r}
gset <- rma(gset)
```
#Quantile normalization
#gset<- normalize(gset, method='quantile')

## Apply log2 transformation if Required
```{r}
ex <- exprs(gset)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
  (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
exprs(gset) <- log2(ex) }
```

## Check probe intensity plot after normalization
```{r}
boxplot(gset, xlab='Samples', ylab='log2Expression',col='Green',
        main= 'Average Expression After Normalization')
hist(gset, lwd=2,xlab='log intensity',
     main="Signal Densities After Normalization")
```

## Naming the Probe IDs to Gene IDs
```{r}
ID<- featureNames(gset)
Symbol <- getSYMBOL(ID,'hgu133plus2.db')
fData(gset) <- data.frame(Symbol=Symbol)
```

## Perform DEG Analysis
```{r}
Groups<- factor(c("Normal", "Normal", "Normal","Normal", "Normal", 
                  "Normal","Tumor","Tumor","Tumor","Tumor","Tumor","Tumor"))
design<- model.matrix(~Groups-1)
colnames(design)<- c("Normal","Tumor")
head(design)
fit<- lmFit(gset, design)
contrast.mat<- makeContrasts("Tumor-Normal", levels = design)
fit2<- contrasts.fit(fit, contrast.mat)
fit3<-eBayes(fit2)
options(digits = 2)
topTable(fit3)
```

```{r}
DEGs<- topTable(fit3, n=Inf, adjust="BH")
write.csv(DEGs, file="DEGs.csv")
```

## Stratify
```{r}
Significant_DEGs<- filter(DEGs,logFC>1|logFC< -1 & adj.P.Val<0.7) 
Significant_UP<- filter(Significant_DEGs,logFC> 1)
Significant_Down<- filter(Significant_DEGs,logFC< -1)
```

## Create basic volcano plot
```{r}
EnhancedVolcano(DEGs,
                lab = DEGs$Symbol,
                x = 'logFC',
                y = 'adj.P.Val',
                pCutoff = 1,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 6.0,
                border='full')
```

## Visualizing DEG of Interest
```{r}
DEG_of_Interest <- data.frame(topTable(fit3, number=Inf,  lfc=1, p.value = 1))
Genes_of_interest <- gset[rownames(DEG_of_Interest),]
heatmap(exprs(Genes_of_interest))

```

##################################################################
## UMAP plot (dimensionality reduction)
```{r}
ex <- na.omit(ex) # eliminate rows with NAs
ex <- ex[!duplicated(ex), ]  # remove duplicates
ump <- umap(t(ex), n_neighbors = 7, random_state = 123)
#par(mar=c(3,3,2,6), xpd=F)
plot(ump$layout, main="UMAP plot, nbrs=7", xlab="", ylab="", col=Groups, pch=20, cex=1.5)
legend('bottomright', inset=c(-0.05,0), legend=levels(Groups), pch=15,
        col=1:nlevels(Groups), title="Group", pt.cex=1.5)
pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)
```

#Saving normalized expression
```{r}
head(exprs(gset),10)
norm.exprs<- data.frame(exprs(gset))
norm.exprs$Symbol <-  getSYMBOL(ID,'hgu133plus2.db')
norm.exprs<-norm.exprs %>% relocate(Symbol)
write.csv(norm.exprs, 'normalized_expression.csv')
```
#################################################################

